Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-11 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is valid; these rotation-based ages are roughly an
  order of magnitude more precise than isochrone ages in this sample
  of stars.
  %
  The independent lithium ages agree with the rotation-based ages for
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler
  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over
  the past 3\,Gyr, equally for both planet hosts and for the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; for most known
exoplanets, most of the action has already happened.

Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
help build the timeline for exoplanet evolution.  Informative
individual young systems include V1298~Tau, an adolescent resonant
chain of four transiting puffy planets that is a likely precusor to
the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
Thao et al.~submitted).  Population-level analyses have similarly
suggested differences in the size distribution of young exoplanets
relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.

@@

@@

\subsection{Future Directions}
{\it Occurrence Rates}:
Some interesting age-evolution trends have been predicted, and can
probably be teased from our data.  Kelvin-Helmholtz cooling for
mini-Neptunes should scale $R_{\rm p}$ as $t^{-0.1}$
\citep{Gupta_2019}.  The photoevaporation desert has been predicted to
be ``carved'' over the timescale for atmospheric loss
\citep{Owen_Lai_2018}.  Mini-Neptunes should be converted into
super-Earths \citep{Rogers_2021}.  And the boundary of the super-Earth
population may correspondingly move up in size over time
\citep{David_2021}.  We're going to split the two papers.

{\it Field star ages from other surveys}:
In this study we focused on deriving rotation and lithium-based ages
for the Kepler planets.  However surveys such as K2, TESS, HAT, and
WASP have had very different selection functions.  Rotation-based age
dating of their discoveries could potentially clarify their youth, and
thereby help clarify the broader issue of their evolution.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
connected to their birth clusters.  Others have not.  A systematic
search for the birth clusters of the youngest ``field stars'' has, to
our knowledge, not been performed.  The ideal approach for such a
search would be to combine spatial and kinematic information from Gaia
with rotation measurements from TESS.  Given that the decoherence time
of young clusters is $\sim$100\,Myr, one would expect that the
youngest ``field stars'' may have young neighbors that are relatively
easy to identify.

>>>>

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our gyrochronology
  model is applicable; these rotation-based ages are over an order of
  magnitude more precise than isochrone ages in this sample of stars.
  %
  The independent lithium ages agree with the rotation-based ages in
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3,Gyr. This trend holds true for both planet hosts and the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10,Myr to $\approx$1,Gyr timescales. For most known
exoplanets, these processes have already run their course.

Young ($<$1\,Gyr) exoplanets are key to building the timeline for
exoplanet evolution.  Informative individual young systems include
V1298~Tau, an adolescent resonant chain of four transiting puffy
planets that is a likely precusor to the compact multiplanet systems
\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.

@@

%FIXME these paragraphs need improvements in how they incorporate
%Lu+21 (meh), but certainly Mathur+2023.

@@

\subsection{Future Directions}

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.

{\it Field star ages from other surveys}:
While this study focused on deriving rotation and lithium-based ages
for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF, and
ATLAS have different selection functions.
Applying rotation-based age dating to their discoveries could clarify
the youth and evolutionary trends of these populations.  Future
missions like PLATO \citep{Rauer14} and Roman Galactic Plane Time
Domain Survey  will further expand the sample of stars with precisely
determined ages, including those known to host planets.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
linked to their birth clusters, while others have not. A systematic
search for the birth clusters of the youngest ``field stars'' could
combine spatial and kinematic information from Gaia with rotation
measurements from TESS. Given the $\sim$100\,Myr decoherence time of
young clusters, the youngest "field stars" may have easily
identifiable young neighbors.
